---
title: "Appendix 2: Selection of Decomposition Method"
subtitle: "Comparison of Das Gupta, Horiuchi, and Stepwise Replacement"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---

## Overview

This appendix justifies the selection of Das Gupta's (1993) symmetric
decomposition method over two alternatives: Horiuchi's continuous
decomposition (Horiuchi, Wilmoth & Pletcher, 2008) and Stepwise
Replacement. All three methods are applied to identical data and their
results are compared across three criteria: symmetry, sensitivity
stability, and interpretability.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| code-fold: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  DemoDecomp,   # Horiuchi and stepwise_replacement
  readxl, dplyr, tidyr, purrr,
  ggplot2, patchwork, scales,
  gt, knitr, sf, stringr
)

# Load custom Das Gupta function (das_gupta2.R: full permutation averaging)
source("R/das_gupta2.R")
```

------------------------------------------------------------------------

## Data preparation

```{r}
#| label: data-load

df <- read_excel("in/POP_UTCI_mun.xlsx")
mapa <- sf::read_sf("in/mapa_ent_small.shp")

df <- df |>
  left_join(
    mapa |> sf::st_drop_geometry() |> select(cve_ent = CVE_ENT, NOMGEO),
    by = "cve_ent"
  ) |>
  select(-region) |>
  rename(region = NOMGEO) |>
  mutate(
    dis_mov   = dis_caminar + dis_banarse,
    dis_otros = dis_ver + dis_oir + dis_recordar + dis_hablar
  )

age5_levels <- c(
  "0-4",  "5-9",  "10-14", "15-19", "20-24",
  "25-29","30-34","35-39", "40-44", "45-49",
  "50-54","55-59","60-64", "65-69", "70-74",
  "75-79","80-84","85+"
)

discapacidades <- c("dis_mov", "dis_mult", "dis_otros")
```

### State-level aggregation

```{r}
#| label: state-data

datos <- list()

for (i in discapacidades) {

  estatal <- df |>
    group_by(region, age5) |>
    summarise(
      S        = sum(pop, na.rm = TRUE),
      D        = sum(.data[[i]], na.rm = TRUE),
      heatdays = weighted.mean(heatdays, w = pop, na.rm = TRUE),
      .groups  = "drop"
    ) |>
    group_by(region) |>
    mutate(s = S / sum(S, na.rm = TRUE)) |>
    ungroup() |>
    mutate(d = D / S, h = heatdays / 366) |>
    group_by(region) |>
    mutate(
      exposure    = sum(d * s * h * 100, na.rm = TRUE),
      specific_exp = d * s * h * 100
    ) |>
    ungroup() |>
    mutate(discapacidad = i)

  nacional <- df |>
    group_by(age5) |>
    summarise(
      S        = sum(pop, na.rm = TRUE),
      D        = sum(.data[[i]], na.rm = TRUE),
      heatdays = weighted.mean(heatdays, w = pop, na.rm = TRUE),
      .groups  = "drop"
    ) |>
    mutate(region = "Nacional") |>
    group_by(region) |>
    mutate(s = S / sum(S, na.rm = TRUE)) |>
    ungroup() |>
    mutate(d = D / S, h = heatdays / 366) |>
    group_by(region) |>
    mutate(
      exposure     = sum(d * s * h * 100, na.rm = TRUE),
      specific_exp = d * s * h * 100
    ) |>
    ungroup() |>
    mutate(discapacidad = i)

  datos[[i]] <- bind_rows(estatal, nacional)
}

datos <- do.call(rbind, datos) |>
  mutate(age5 = factor(age5, levels = age5_levels))

# Datos para descomposición (combined across disability types for joint exposure)
datos_decomp <- datos |>
  group_by(region, age5) |>
  mutate(exposure = sum(specific_exp, na.rm = TRUE)) |>
  ungroup()

# Reference and comparison regions
ref     <- "Nacional"
regiones <- setdiff(unique(datos_decomp$region), ref)

# Real differences
valores_reales <- datos_decomp |>
  group_by(region) |>
  summarise(
    exposure = sum(d * s * h * 100, na.rm = TRUE),
    .groups  = "drop"
  ) |>
  mutate(
    ref_exposure  = exposure[region == ref],
    diff_real     = exposure - ref_exposure,
    diffprop_real = diff_real / ref_exposure * 100
  )
```

------------------------------------------------------------------------

## Exposure function

All three methods operate on the same exposure function:

$$E = 100 \times \sum_i s_i \times d_i \times h_i$$

```{r}
#| label: exposure-fun
#| code-fold: false

exposure_fun <- function(pars) {
  n <- length(pars) / 3
  d <- pars[1:n]
  s <- pars[(n + 1):(2 * n)]
  h <- pars[(2 * n + 1):(3 * n)]
  sum(d * s * h * 100)
}

# Parameter vectors by region (order: d, s, h)
data_list  <- split(datos_decomp, datos_decomp$region)
parameters <- lapply(data_list, function(df) c(df$d, df$s, df$h))

# Parameter names for labelling contributions
ids <- datos_decomp |>
  filter(region == ref) |>
  mutate(id = paste0(age5, "__", discapacidad)) |>
  pull(id)

param_names <- c(
  paste0("d__", ids),
  paste0("s__", ids),
  paste0("h__", ids)
)
```

------------------------------------------------------------------------

## Method 1: Horiuchi continuous decomposition

```{r}
#| label: horiuchi
#| cache: true

# N = 40 eliminates numerical ambiguity across all Mexican regions
N_horiuchi <- 40

horiuchi_list <- list()
for (region in regiones) {
  horiuchi_list[[region]] <- DemoDecomp::horiuchi(
    func  = exposure_fun,
    pars1 = parameters[[ref]],
    pars2 = parameters[[region]],
    N     = N_horiuchi
  )
  names(horiuchi_list[[region]]) <- param_names
}

horiuchi_df <- purrr::imap_dfr(
  horiuchi_list,
  ~ tibble(region = .y, parameter = names(.x), contribution = as.numeric(.x))
) |>
  tidyr::separate(parameter,
                  into = c("component", "age5", "discapacidad"),
                  sep = "__", convert = TRUE) |>
  mutate(age5 = factor(age5, levels = age5_levels))

cat("✓ Horiuchi complete  (N =", N_horiuchi, ")\n")
```

## Method 2: Das Gupta symmetric decomposition

```{r}
#| label: dasgupta
#| cache: true

dasgupta_list <- purrr::map(regiones, function(reg) {
  res <- das_gupta(
    df          = datos_decomp,
    ref         = ref,
    cmp         = reg,
    region_var  = region,
    strata_vars = c("age5", "discapacidad"),
    comp_vars   = c("d", "s", "h"),
    scale       = 100
  )
  res$global
})

dasgupta_df <- bind_rows(dasgupta_list) |>
  rename(
    Horiuchi_ref   = R_ref,     # solo para no confundir con horiuchi_df
    effect_d       = effect_d,
    effect_s       = effect_s,
    effect_h       = effect_h
  )

cat("✓ Das Gupta complete\n")
```

## Method 3: Stepwise Replacement

```{r}
#| label: stepwise
#| cache: true

stepwise_list <- list()
for (region in regiones) {
  stepwise_list[[region]] <- DemoDecomp::stepwise_replacement(
    func  = exposure_fun,
    pars1 = parameters[[ref]],
    pars2 = parameters[[region]]
  )
  names(stepwise_list[[region]]) <- param_names
}

stepwise_df <- purrr::imap_dfr(
  stepwise_list,
  ~ tibble(region = .y, parameter = names(.x), contribution = as.numeric(.x))
) |>
  tidyr::separate(parameter,
                  into = c("component", "age5", "discapacidad"),
                  sep = "__", convert = TRUE) |>
  mutate(age5 = factor(age5, levels = age5_levels))

cat("✓ Stepwise complete\n")
```

------------------------------------------------------------------------

## Comparison

### Aggregate contributions by region and component

```{r}
#| label: aggregate-contributions

horiuchi_totales  <- horiuchi_df |>
  group_by(region, component) |>
  summarise(Horiuchi = sum(contribution, na.rm = TRUE), .groups = "drop")

dasgupta_totales <- dasgupta_df |>
  select(region, d = effect_d, s = effect_s, h = effect_h) |>
  pivot_longer(
    cols      = c(d, s, h),
    names_to  = "component",
    values_to = "Das Gupta"
  )

stepwise_totales  <- stepwise_df |>
  group_by(region, component) |>
  summarise(Stepwise = sum(contribution, na.rm = TRUE), .groups = "drop")

# Join all methods
# NOTE: pivot_wider produces "Das Gupta" (with space); use backtick references
todos_metodos <- horiuchi_totales |>
  left_join(dasgupta_totales,  by = c("region", "component")) |>
  left_join(stepwise_totales,  by = c("region", "component")) |>
  left_join(valores_reales |> select(region, diff_real, diffprop_real),
            by = "region")
```

### Additivity check

Each method must satisfy: $C(s) + C(d) + C(h) = \Delta E$

```{r}
#| label: additivity-check

todos_metodos |>
  group_by(region) |>
  summarise(
    sum_Horiuchi  = sum(Horiuchi,    na.rm = TRUE),
    sum_DasGupta  = sum(`Das Gupta`, na.rm = TRUE),
    sum_Stepwise  = sum(Stepwise,    na.rm = TRUE),
    diff_real     = unique(diff_real)
  ) |>
  mutate(
    add_Horiuchi = abs(sum_Horiuchi  - diff_real) < 1e-6,
    add_DasGupta = abs(sum_DasGupta  - diff_real) < 1e-6,
    add_Stepwise = abs(sum_Stepwise  - diff_real) < 1e-6
  ) |>
  summarise(
    across(starts_with("add_"), ~ mean(.x, na.rm = TRUE),
           .names = "{.col}_pct_exact")
  ) |>
  gt() |>
  tab_header(title = "Additivity check: proportion of states with exact decomposition")
```

### Consensus distance (Table A2)

Since the true per-component contribution is unobservable, we use the
median across the three methods as a consensus reference.

```{r}
#| label: consensus-table

comparacion_consenso <- todos_metodos |>
  rowwise() |>
  mutate(
    consenso       = median(c(Horiuchi, `Das Gupta`, Stepwise), na.rm = TRUE),
    err_Horiuchi   = abs(Horiuchi    - consenso),
    err_DasGupta   = abs(`Das Gupta` - consenso),
    err_Stepwise   = abs(Stepwise    - consenso)
  ) |>
  ungroup()

# Table A2: mean error by component
tabla_a2 <- comparacion_consenso |>
  group_by(component) |>
  summarise(
    `Das Gupta` = mean(err_DasGupta,  na.rm = TRUE),
    Horiuchi    = mean(err_Horiuchi,  na.rm = TRUE),
    Stepwise    = mean(err_Stepwise,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(component = case_when(
    component == "s" ~ "Age structure (s)",
    component == "d" ~ "Disability prevalence (d)",
    component == "h" ~ "Heat exposure (h)"
  ))

tabla_a2 |>
  gt() |>
  tab_header(
    title    = "Table A2. Mean distance to consensus value by method and component",
    subtitle = "Consensus = median contribution across the three methods"
  ) |>
  fmt_number(columns = c(`Das Gupta`, Horiuchi, Stepwise), decimals = 10) |>
  tab_footnote(
    "The consensus value represents the median contribution estimate across the
     three methods for each component and region. Since the true component-specific
     contribution is unobservable (only the total difference is observed), we use
     the median as a reference point to evaluate which method produces estimates
     closest to the central tendency across all three approaches."
  ) |>
  cols_label(component = "Component")
```

### Sample comparison table (Table A1)

```{r}
#| label: sample-table

sample_states <- c("Aguascalientes", "Baja California", "Oaxaca",
                   "Ciudad de México", "Nuevo León")

todos_metodos |>
  filter(region %in% sample_states) |>
  mutate(component = case_when(
    component == "d" ~ "d",
    component == "s" ~ "s",
    component == "h" ~ "h"
  )) |>
  arrange(region, component) |>
  select(region, component, Horiuchi, `Das Gupta`, Stepwise) |>
  gt() |>
  tab_header(
    title    = "Table A1. Decomposition results for selected states",
    subtitle = "Contribution of each component by method"
  ) |>
  fmt_number(columns = c(Horiuchi, `Das Gupta`, Stepwise), decimals = 4) |>
  cols_label(
    region    = "State",
    component = "Component"
  )
```

------------------------------------------------------------------------

## Figure: Method comparison by component

```{r}
#| label: fig-method-comparison
#| fig-width: 14
#| fig-height: 5
#| fig-cap: "Figure A2. Comparison of component contributions across decomposition methods"

comparacion_long <- todos_metodos |>
  pivot_longer(
    cols      = c(Horiuchi, `Das Gupta`, Stepwise),
    names_to  = "method",
    values_to = "contribution"
  ) |>
  mutate(
    method    = factor(method, levels = c("Horiuchi", "Das Gupta", "Stepwise")),
    component = case_when(
      component == "s" ~ "Age structure (s)",
      component == "d" ~ "Disability prevalence (d)",
      component == "h" ~ "Heat exposure (h)"
    )
  )

ggplot(comparacion_long,
       aes(x = contribution, y = fct_reorder(region, contribution),
           color = method, shape = method)) +
  geom_point(alpha = 0.75, size = 1.8) +
  facet_wrap(~ component, scales = "free_x", ncol = 3) +
  scale_color_manual(
    values = c("Horiuchi" = "#2E86AB", "Das Gupta" = "#E84855", "Stepwise" = "#3BB273")
  ) +
  labs(
    x = "Contribution to difference from national average",
    y = NULL,
    color = "Method", shape = "Method"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(face = "bold"),
    axis.text.y     = element_text(size = 7)
  )
```

------------------------------------------------------------------------

## Method selection rationale

Based on the comparison above, Das Gupta's method is selected for three
reasons:

**1. Order independence.** By averaging across all 3! = 6 possible
replacement sequences (sdh, shd, dsh, dhs, hsd, hds), it eliminates
arbitrary decisions about which component to replace first. There is no
theoretical justification for prioritizing age structure, disability
prevalence, or heat exposure.

**2. Demonstrated stability.** Table A2 shows that Das Gupta produces
the estimates closest to the consensus value, particularly for the
disability prevalence component (d), where interaction effects with the
other two components are strongest.

**3. Clear interpretation.** Each component receives its "average
contribution" to the total difference — a conceptually transparent
interpretation that is standard in demographic analysis (Das Gupta,
1993; Preston et al., 2001).

### Acknowledged limitations

-   **Computational cost:** Grows factorially with the number of
    components (manageable here: 3! = 6 permutations).
-   **Custom implementation required:** `DemoDecomp` does not export a
    `das_gupta()` function; the custom function in `R/das_gupta2.R` is
    used throughout this study.
-   **Implicit interactions:** Like all symmetric methods, Das Gupta
    distributes interaction terms across components without explicitly
    modelling them. This is a deliberate design choice rather than a
    limitation when no hierarchical ordering is theoretically justified.
