---
title: "Appendix 1: Municipal vs. State-Level Decomposition"
subtitle: "Sensitivity Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---

## Overview

State-level aggregation was adopted for practical and policy-relevant
reasons. To assess whether this choice masks important spatial patterns,
we compare two analytical strategies:

-   **Approach 1:** Perform Das Gupta decomposition at the municipal
    level, then aggregate component contributions to the state level
    using population weights.
-   **Approach 2:** Aggregate demographic and climatic data to the state
    level first, then perform decomposition.

If both approaches yield highly similar component estimates, state-level
decomposition is a valid and efficient analytical choice.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| code-fold: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  readxl, dplyr, tidyr, purrr,
  ggplot2, patchwork, scales,
  gt, knitr, sf
)

# Custom functions
source("R/das_gupta.R")            # Simple Das Gupta (one region vs reference)
source("R/prepare_horiuchi_data.R") # State-level data aggregation
```

------------------------------------------------------------------------

## Data preparation

```{r}
#| label: data-load

df <- read_excel("in/POP_UTCI_mun.xlsx")
mapa <- sf::read_sf("in/mapa_ent_small.shp")

df <- df |>
  left_join(
    mapa |> sf::st_drop_geometry() |> select(cve_ent = CVE_ENT, NOMGEO),
    by = "cve_ent"
  ) |>
  select(-region) |>
  rename(region = NOMGEO) |>
  mutate(
    dis_mov   = dis_caminar + dis_banarse,
    dis_otros = dis_ver + dis_oir + dis_recordar + dis_hablar
  )

age5_levels <- c(
  "0-4",  "5-9",  "10-14", "15-19", "20-24",
  "25-29","30-34","35-39", "40-44", "45-49",
  "50-54","55-59","60-64", "65-69", "70-74",
  "75-79","80-84","85+"
)

# Three disability subtypes (popdis excluded from sensitivity comparison
# to focus on subtypes with clearest policy interpretation)
discapacidades <- c("dis_mov", "dis_mult", "dis_otros")
```

### Approach 2: State-level data

For each state $e$, disability type $k$, and age group $i$, the three
components are constructed as:

$$
S_{e,k,i} = \sum_m POP_{m,k,i}, \quad
D_{e,k,i} = \sum_m POP\_DIS_{m,k,i}, \quad
H_{e,k,i} = \frac{\sum_m POP_{m,k,i} \cdot HeatDays_m}{\sum_m POP_{m,k,i}}
$$

```{r}
#| label: state-data

# prepare_horiuchi_data() aggregates df to state level for each
# disability type, computing s, d, h and exposure identically
# to the main analysis pipeline
datos_estatal <- prepare_horiuchi_data(df, discapacidades, age5_levels)

cat("States prepared:", n_distinct(datos_estatal$region) - 1, "\n")
cat("Disability types:", n_distinct(datos_estatal$discapacidad), "\n")
```

### Approach 1: Municipal-level data

```{r}
#| label: municipal-data

datos_municipal <- list()

for (i in discapacidades) {

  municipal <- df |>
    group_by(cve_ent, cvegeo, region, age5) |>
    summarise(
      S        = sum(pop, na.rm = TRUE),
      D        = sum(.data[[i]], na.rm = TRUE),
      heatdays = weighted.mean(heatdays, w = pop, na.rm = TRUE),
      .groups  = "drop"
    ) |>
    group_by(cve_ent, cvegeo, region) |>
    mutate(s = S / sum(S, na.rm = TRUE)) |>
    ungroup() |>
    mutate(d = D / S, h = heatdays / 366) |>
    group_by(cve_ent, cvegeo, region) |>
    mutate(exposure = sum(d * s * h * 100, na.rm = TRUE)) |>
    ungroup() |>
    mutate(discapacidad = i)

  nacional <- df |>
    group_by(age5) |>
    summarise(
      S        = sum(pop, na.rm = TRUE),
      D        = sum(.data[[i]], na.rm = TRUE),
      heatdays = weighted.mean(heatdays, w = pop, na.rm = TRUE),
      .groups  = "drop"
    ) |>
    mutate(region = "Nacional", cve_ent = "00", cvegeo = "00000") |>
    group_by(region) |>
    mutate(s = S / sum(S, na.rm = TRUE)) |>
    ungroup() |>
    mutate(d = D / S, h = heatdays / 366) |>
    group_by(region) |>
    mutate(exposure = sum(d * s * h * 100, na.rm = TRUE)) |>
    ungroup() |>
    mutate(discapacidad = i)

  datos_municipal[[i]] <- bind_rows(municipal, nacional)
}

datos_municipal <- do.call(rbind, datos_municipal) |>
  mutate(age5 = factor(age5, levels = age5_levels))

cat("Municipalities:", n_distinct(datos_municipal$cvegeo) - 1, "\n")
```

------------------------------------------------------------------------

## Decomposition

### Approach 2: Direct state-level Das Gupta

```{r}
#| label: decomp-state

resultados_estatal <- list()

for (dis_type in discapacidades) {

  data_decomp <- datos_estatal |> filter(discapacidad == dis_type)
  regions     <- setdiff(unique(data_decomp$region), "Nacional")

  results_dis <- purrr::map_dfr(
    regions,
    ~ das_gupta(df = data_decomp, ref_region = "Nacional", cmp_region = .x)
  ) |> mutate(discapacidad = dis_type)

  resultados_estatal[[dis_type]] <- results_dis
  cat("✓", dis_type, "-", nrow(results_dis), "states\n")
}

resultados_estatal_final <- do.call(rbind, resultados_estatal)
```

### Approach 1: Municipal Das Gupta + state aggregation

```{r}
#| label: decomp-municipal
#| cache: true

resultados_municipal <- list()

for (dis_type in discapacidades) {

  cat("\nProcessing:", dis_type, "\n")
  data_decomp <- datos_municipal |> filter(discapacidad == dis_type)
  municipios  <- data_decomp |>
    filter(region != "Nacional") |>
    distinct(cvegeo, region, cve_ent)

  pb <- txtProgressBar(min = 0, max = nrow(municipios), style = 3)
  results_list <- list()

  for (i in seq_len(nrow(municipios))) {
    mun_info  <- municipios[i, ]
    data_temp <- data_decomp |>
      filter(cvegeo == mun_info$cvegeo | region == "Nacional") |>
      mutate(region = if_else(
        cvegeo == mun_info$cvegeo,
        paste0("MUN_", cvegeo),
        "Nacional"
      ))

    result <- tryCatch(
      das_gupta(
        df         = data_temp,
        ref_region = "Nacional",
        cmp_region = paste0("MUN_", mun_info$cvegeo)
      ) |> mutate(
        cvegeo  = mun_info$cvegeo,
        cve_ent = mun_info$cve_ent,
        region  = mun_info$region
      ),
      error = function(e) NULL
    )

    results_list[[i]] <- result
    setTxtProgressBar(pb, i)
  }

  close(pb)

  resultados_municipal[[dis_type]] <- bind_rows(results_list) |>
    mutate(discapacidad = dis_type)

  cat("\n  ✓ Done:", nrow(resultados_municipal[[dis_type]]), "municipalities\n")
}

resultados_municipal_final <- do.call(rbind, resultados_municipal)
```

### Aggregate municipal results to state level

```{r}
#| label: aggregate-to-state
#| message: false
#| warning: false
#| results: hide

# Join population weights from municipal data
pop_weights <- datos_municipal |>
  filter(region != "Nacional") |>
  group_by(cvegeo, cve_ent, discapacidad) |>
  summarise(pop_total = sum(S, na.rm = TRUE), .groups = "drop")

# Weighted aggregation to state
resultados_mun_agg <- resultados_municipal_final |>
  left_join(pop_weights, by = c("cvegeo", "cve_ent", "discapacidad")) |>
  group_by(region, discapacidad) |>
  summarise(
    effect_age_mun        = weighted.mean(effect_age,        pop_total, na.rm = TRUE),
    effect_disability_mun = weighted.mean(effect_disability, pop_total, na.rm = TRUE),
    effect_heat_mun       = weighted.mean(effect_heat,       pop_total, na.rm = TRUE),
    n_municipios          = n(),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

## Comparison

```{r}
#| label: comparison

comparacion <- resultados_estatal_final |>
  select(region, discapacidad,
         effect_age_dir        = effect_age,
         effect_disability_dir = effect_disability,
         effect_heat_dir       = effect_heat,
         total_diff) |>
  left_join(resultados_mun_agg, by = c("region", "discapacidad")) |>
  mutate(
    diff_age        = effect_age_dir        - effect_age_mun,
    diff_disability = effect_disability_dir - effect_disability_mun,
    diff_heat       = effect_heat_dir       - effect_heat_mun
  )
```

### Pearson correlations

```{r}
#| label: correlations

cor_age        <- cor(comparacion$effect_age_mun,
                      comparacion$effect_age_dir,        use = "complete.obs")
cor_disability <- cor(comparacion$effect_disability_mun,
                      comparacion$effect_disability_dir, use = "complete.obs")
cor_heat       <- cor(comparacion$effect_heat_mun,
                      comparacion$effect_heat_dir,       use = "complete.obs")

tibble(
  Component      = c("Age structure (s)", "Disability prevalence (d)", "Heat exposure (h)"),
  `Pearson r`    = round(c(cor_age, cor_disability, cor_heat), 4),
  Interpretation = case_when(
    c(cor_age, cor_disability, cor_heat) >= 0.99 ~ "Near-perfect agreement",
    c(cor_age, cor_disability, cor_heat) >= 0.95 ~ "Very strong agreement",
    TRUE ~ "Strong agreement"
  )
) |>
  gt() |>
  tab_header(
    title    = "Table A1. Pearson Correlations between Approaches",
    subtitle = "Municipal aggregation vs. direct state-level decomposition"
  ) |>
  fmt_number(columns = `Pearson r`, decimals = 4) |>
  cols_align(align = "center", columns = `Pearson r`) |>
  tab_footnote("N = 32 states × 3 disability types. Pearson r computed with complete observations.")
```

### Figure A1: Scatter plots by component

```{r}
#| label: fig-scatter
#| fig-width: 12
#| fig-height: 5
#| fig-cap: "Figure A1. Sensitivity analysis: municipal aggregation vs. direct state-level decomposition"

plot_component <- function(data, x_var, y_var, label, r_val) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]],
                   color = discapacidad)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "gray40", linewidth = 0.8) +
    geom_point(alpha = 0.7, size = 2.2) +
    scale_color_manual(
      values = c("dis_mov" = "#E07B54", "dis_mult" = "#5B9BD5", "dis_otros" = "#70AD47"),
      labels = c("Mobility", "Multiple", "Other"),
      name   = "Disability type"
    ) +
    labs(
      title    = label,
      x        = "Municipal aggregation",
      y        = "Direct state decomposition",
      subtitle = sprintf("Pearson correlation: r = %.4f  |  N = %d states × 3 disability types",
                         r_val, 32)
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom",
          plot.title = element_text(face = "bold", size = 12))
}

p1 <- plot_component(comparacion, "effect_age_mun",        "effect_age_dir",
                     "Age Structure Component (s)", cor_age)
p2 <- plot_component(comparacion, "effect_disability_mun", "effect_disability_dir",
                     "Disability Prevalence Component (d)", cor_disability)
p3 <- plot_component(comparacion, "effect_heat_mun",       "effect_heat_dir",
                     "Heat Exposure Component (h)", cor_heat)

fig_a1 <- (p1 | p2 | p3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave("out/fig_A1_sensitivity_scatter.png",
       plot   = fig_a1,
       width  = 12,
       height = 5,
       dpi    = 300)

fig_a1
```

------------------------------------------------------------------------

## Interpretation

The diagonal dashed line in each panel represents perfect agreement
between methods. Points falling close to the diagonal indicate that both
approaches yield similar component estimates for a given state and
disability type.

Pearson correlations of r = `r round(cor_age, 2)` for age structure, r =
`r round(cor_disability, 2)` for disability prevalence, and r =
`r round(cor_heat, 2)` for heat exposure confirm that state-level
decomposition preserves the substantive territorial patterns identified
by the more computationally intensive municipal approach. State-level
aggregation is therefore a valid and efficient analytical strategy for
the purposes of this study.
